# Reproducible Research: Peer Assessment 1
==========================================

```{r setparam, echo = FALSE}
opts_chunk$set(echo = TRUE)
```
## Loading and preprocessing the data
```{r}
data_file = read.csv("./activity.csv", stringsAsFactors = FALSE, header = TRUE, na.string = "NA")
data_file$date = as.Date(data_file$date)
```

## What is mean total number of steps taken per day?
```{r meanandmedian}
library(reshape)
datamelt = melt(data_file, id = c("date"), measure.vars = c("steps"))
steps_count = cast(datamelt, date ~ variable, sum)
```

- Plot the Histogram
```{r plothistogram, fig.height = 8, fig.width = 16}
plot(steps_count, type = "h", lwd = 20)
```

- Calculate and report the mean steps taken every day
```{r meansteps}
meansteps = cast(datamelt, date ~ variable, mean, na.rm = TRUE)
names(meansteps) = c("date", "mean_steps")
print(meansteps)
```

- Calculate and report the median steps taken every day
```{r medianvalue}
mediansteps = cast(datamelt, date ~ variable, median, na.rm = TRUE)
names(mediansteps) = c("date", "median_steps")
print(mediansteps)
```

## What is the average daily activity pattern?
```{r}
datamelt2 = melt(data_file, id = c("interval"), measure.vars = c("steps"))
average_steps = cast(datamelt2, interval ~ variable, mean, na.rm = TRUE)
names(average_steps) = c("interval", "avsteps")
```

- Plot the timeseries
```{r plottimeseries, fig.height = 8, fig.width = 16}
plot(average_steps, type = "l")
```

- Calculate max # of steps for an interval across all days 
```{r caclmax}
max_value = max(average_steps$avsteps)
interval_number = average_steps[average_steps$avsteps == max_value, "interval"]
cat("The interval with max number of steps averaged across all days is = ", 
    interval_number)
```
## Imputing missing values

-Missing Values
```{r missingvalues}
num_complete_cases = sum(!complete.cases(data_file))
cat("No. of missing values is=", num_complete_cases)
```
- Create new dataset with missing values filled in: The strategy is to fill up the missing values with the mean for that 5-minute interval. The code iterates over the entire dataset. Wherever the # of steps is NA, it fills it up with the mean value of the steps taken in that 5-minute interval
```{r createnewdataset}
new_data_file = data_file
for (i in 1:nrow(data_file)) {
    if (is.na(data_file[i, "steps"])) {
        key_interval = data_file[i, "interval"]
        new_data_file[i, "steps"] = average_steps[average_steps$interval == 
            key_interval, "avsteps"]
    }
}
```

- Histogram of NEW total # of steps taken each day
```{r newtotalsteps, fig.height = 8, fig.width = 16}
new_datamelt = melt(new_data_file, id = c("date"), measure.vars = c("steps"))
steps_count = cast(new_datamelt, date ~ variable, sum)
plot(steps_count, type = "h", lwd = 20)
```

- Calculate the new mean steps taken per day
```{r newmean}
new_meansteps = cast(new_datamelt, date ~ variable, mean, na.rm = TRUE)
names(new_meansteps) = c("date", "mean_steps")
print(new_meansteps)
```

- Calculate the new median steps taken per day
```{r newmedian}
new_mediansteps = cast(new_datamelt, date ~ variable, median, na.rm = TRUE)
names(new_mediansteps) = c("date", "median_steps")
print(new_mediansteps)
```

## Are there differences in activity patterns between weekdays and weekends?
- Create a new factor variable in the dataset with levels - weekday and weekend
```{r createfactorvariable}
new_data_file$weekdays = ifelse(weekdays(new_data_file$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
```
- Calculate the mean # of steps over weekday and weekend
```{r}
datamelt3 = melt(new_data_file, id = c("interval", "weekdays"), measure.vars = c("steps"))
avg_steps = cast(datamelt3, interval + weekdays ~ variable, mean)
```
- Plot the mean steps for weekend and weekday
```{r plotweekdaysmean, fig.height= 8, fig.width = 16}
library(lattice)
xyplot(avg_steps$steps ~ avg_steps$interval | avg_steps$weekdays, data = avg_steps, layout = c(1, 2), type = "l", xlab = "Interval", ylab = "Number of Steps")
```